---
title: "Reproducible Research: Peer Assessment 2"
output: 
  html_document:
    keep_md: true
---

# Title: JHU-DSCI Reproducible Research
## Peer Assessment 2 - NOAA Storm Data

### Name: Walter Yu
### Date: July 2020

## Synopsis

This document completes an initial analysis of the NOAA Storm Data set as follows:  
1. Download and read the raw NOAA storm dataset 
2. Review and clean data before analysis 
3. Analyze data to answer assignment questions 
4. Plot data to communicate results 
5. Document steps and findings to be reproducible

## Submission Notes

1. This markdown file is an analysis and visualization of the NOAA weather dataset. Source data available [here][1].

[1]: https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2

2. The document for this assignment was generated with [R Studio Cloud][2] due to package installation errors on my local machine; as a result, file was run and report rendered there instead.

[2]: https://rstudio.cloud

This markdown project template is based on a fork of the course assignment Github repo available [here][3].

[3]: https://github.com/rdpeng/RepData_PeerAssessment1

This assignment is completed for the JHU Coursera Data Science Program, which is a 10-course certification. More info about this program is available [here][4].

[4]: https://www.coursera.org/specializations/jhu-data-science

## Questions

1. Across the United States, which types of events (as indicated in the EVTYPE variable) are most harmful with respect to population health?

2. Across the United States, which types of events have the greatest economic consequences?

## Part 1: Data Processing

### Part 1A: Data Import

Steps:
1. Loaded data with read.csv function per instructions
2. File was unzipped, then read into program

### Part 1B: Data Clean

Processing Steps:
1. Used na.omit function to remove null values
2. Assigned results to a new variable

Analysis:
1. Used head, names and summary functions to review the data

```{r echo=TRUE}

# part 1A: import data per assignment instructions

# source: https://rpubs.com/otienodominic/398952
url <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
temp <- tempfile()
download.file(url, temp)
data <- read.csv(bzfile("repdata_data_StormData.csv.bz2"))
unlink(zipfile)

# source: assignment instructions
# data <- read.csv("data/repdata_data_StormData.csv.bz2")

# review dataset
# https://www.statmethods.net/stats/descriptives.html
head(data)
names(data)
summary(data)

# part 2: remove null values
# source: https://www.statmethods.net/input/missingdata.html
# activity_omit <- na.omit(activity)

# verify results
# https://www.statmethods.net/stats/descriptives.html
# head(activity_omit)
# names(activity_omit)
# summary(activity_omit)

```

## What is mean total number of steps taken per day?

### Part 1: Calculate sum, then create histogram 

Processing Steps:
* Used aggreage function to calculate total steps/day
* Assigned results to a new variable

Analysis:
* Used head, names and summary functions to review the data
* Verified that aggregate returned total steps per day

Plot:
* Created histogram for steps/day; axis adjusted to fit data
* Applied bins to visualize the data effectively

### Part 2: Calculate mean/median by date

Processing Steps:
* Used aggreage function to calculate mean/median steps/day
* Assigned results to a new variable

Analysis:
* Used head, names and summary functions to review the data
* Verified that aggregate returned mean/median steps per day
* Median steps/day = 0 due to large number of zero step values

```{r echo=TRUE}

# part 1: calculate sum, then create histogram 
# aggregate by date, then apply calculations
# source: https://www.statmethods.net/management/aggregate.html
steps_sum <- aggregate(steps ~ date, activity_omit, sum)

# verify results
# https://www.statmethods.net/stats/descriptives.html
head(steps_sum)
names(steps_sum)
summary(steps_sum)

# create histogram plot; adjust axis and bins
# source: https://www.statmethods.net/graphs/density.html
hist(steps_sum$steps, col="green", 
     breaks=10, main="Total Steps per Day",
     xlab="Steps per Day", ylim=c(0,20), xlim=c(0,25000))

# part 2: calculate mean/median by date
# aggregate by date, then apply calculations
# source: https://www.statmethods.net/management/aggregate.html
steps_mean <- aggregate(steps ~ date, activity_omit, mean)

# verify results
# https://www.statmethods.net/stats/descriptives.html
head(steps_mean)
names(steps_mean)
summary(steps_mean)

# aggregate by date, then apply calculations
# source: https://www.statmethods.net/management/aggregate.html
steps_median <- aggregate(steps ~ date, activity_omit, median)

# verify results
# https://www.statmethods.net/stats/descriptives.html
head(steps_median)
names(steps_median)
summary(steps_median)

```

## What is the average daily activity pattern?

### Part 1: Group by interval, calculate mean and plot time series

Processing Steps:
* Used dplyr/group_by to calculate mean steps/interval
* Assigned results to a new variable

Analysis:
* Used head, names and summary functions to review the data
* Verified that mean of steps/interval differ from steps/day

Plot:
* Created time series plot; add axis labels and line color

Notes:
* Assumed instructions were for steps/interval across ALL DAYS
* So, mean was calculated per interval for entire data set
* NOT group by interval, then calculate mean by date

### Part 2: Find interval with maximum steps across all days

Processing Steps:
* Used which function to subset for maximum steps/interval
* Assigned results to a new variable

Analysis:
* Interval 835 had the maximum steps/interval.

```{r echo=TRUE}

# part 1: group by interval, calculate mean and plot time series
# group by interval, then calculate mean steps for each group 
# source: https://datacarpentry.org/R-genomics/04-dplyr.html
install.packages("dplyr")
library(dplyr)
steps_avg_interval <- activity_omit %>%
    group_by(interval) %>%
    summarize(steps_avg = mean(steps))

# verify results
# https://www.statmethods.net/stats/descriptives.html
head(steps_avg_interval)
names(steps_avg_interval)
summary(steps_avg_interval)

# create time series plot; use date list and sum of steps
# source: https://www.datamentor.io/r-programming/plot-function/
plot(steps_avg_interval,
     main="Average Steps per Interval",
     xlab="Interval",
     ylab="Average Steps",
     col="blue",
     type="l")

# part 2: find interval with maximum steps across all days
# source: https://stackoverflow.com/questions/19449615/how-to-extract-the-row-with-min-or-max-values
max_interval <- steps_avg_interval[which.max(steps_avg_interval$steps_avg),]

# return row with max interval
print("Subset for maximum interval and return result.")
max_interval

```

## Imputing missing values

### Part 1: Return count of rows with null values

Processing Steps:
* Used is.na function to subset for records with null 
* Assigned results to a new variable

Analysis:
* Verified that row count (2308) was same as raw data/summary

### Part 2-3: Create new dataset with imputed null values 

Processing Steps:
* Used simputation/impute_lm to impute null values
* Replaced null values with linear model per interval
* Assigned dataset to new variable

Analysis:
* Imputation did not change mean/median 
* Change did not occur since linear model filled matching values
* However, total steps increased since filled values add to total

Notes:
* Numerous attempts were made with dplyr/group_by which failed
* Code left to document previous attempts
* As a result, simputation package was used to yield result

### Part 4: calculate sum, then create histogram 

Processing Steps:
* Used aggreage function to calculate total steps/day
* Assigned results to a new variable

Plot:
* Created histogram; add axis labels and line color
* Total steps increased since filled values add to total

Analysis:
* Imputation did not change mean/median 
* Change did not occur since linear model filled matching values
* However, total steps increased since filled values add to total

```{r echo=TRUE}

# part 1: return count of rows with null values
# https://stackoverflow.com/questions/7980622/subset-of-rows-containing-na-missing-values-in-a-chosen-column-of-a-data-frame
activity_na <- activity[is.na(activity),]

# return number of records with null values
print("Subset for records with null values and return count.")
nrow(activity_na)

# source: https://cran.r-project.org/web/packages/simputation/vignettes/intro.html
install.packages('simputation')
library(simputation)
activity_imp <- impute_lm(activity, steps ~ interval)

# verify results
# https://www.statmethods.net/stats/descriptives.html
nrow(activity_imp)
head(activity_imp)
names(activity_imp)

# compare raw and imputed data
# https://www.statmethods.net/stats/descriptives.html
summary(activity)
summary(activity_imp)

# part 4: calculate sum, then create histogram 
# aggregate by date, then apply calculations
# source: https://www.statmethods.net/management/aggregate.html
steps_sum_imp <- aggregate(steps ~ date, activity_imp, sum)
summary(steps_sum_imp)

# create histogram plot
# source: https://www.statmethods.net/graphs/density.html
hist(steps_sum_imp$steps, col="green", 
     breaks=10, main="Total Steps per Day (Imputed Values)",
     xlab="Steps per Day", ylim=c(0,30), xlim=c(0,25000))

# aggregate by date, then apply calculations
# source: https://www.statmethods.net/management/aggregate.html
steps_mean_imp <- aggregate(steps ~ date, activity_imp, mean)
summary(steps_mean)
summary(steps_mean_imp)

# aggregate by date, then apply calculations
# source: https://www.statmethods.net/management/aggregate.html
steps_median_imp <- aggregate(steps ~ date, activity_imp, median)
summary(steps_median)
summary(steps_median_imp)

```

## Are there differences in activity patterns between weekdays and weekends?

### Part 1: Create weekday/weekend factor within dataset

Processing Steps:
* Created weekday/weekend factor with ifelse logic
* Used weekdays function to create weekday/weekend values

Analysis:
* Used head, names and summary functions to review the data
* Verified that factors were created successfully

### Part 2: Create time series facet plot

Processing Steps:
* Used dplyr to group by date, then calculate mean 

Analysis:
* Used head, names and summary functions to review the data
* Verified that mean was calculated by date

Plot:
* Created time series facet plot; add line color

Notes:
* Assumed instructions were for steps/interval across ALL DAYS
* So, mean was calculated per interval for entire data set
* NOT group by interval, then calculate mean by date

```{r echo=TRUE}

# part 1: create weekday/weekend factor within dataset

# first, create weekday/weekend categorical variable
# source: https://sphweb.bumc.bu.edu/otlt/MPH-Modules/BS/R/R-Manual/R-Manual4.html
activity_imp$wday_type <- ifelse(
    weekdays(as.Date(activity_imp$date)) == "Saturday" |
    weekdays(as.Date(activity_imp$date)) == "Sunday",
    "Weekend",
    "Weekday"
)

# verify results
# https://www.statmethods.net/stats/descriptives.html
nrow(activity_imp)
head(activity_imp)
names(activity_imp)
summary(activity_imp)
unique(activity_imp$wday_type)

# group by interval, then calculate mean steps for each group 
# source: https://datacarpentry.org/R-genomics/04-dplyr.html
# install.packages("dplyr")
# library(dplyr)
steps_avg_wday <- activity_imp %>%
    group_by(steps, interval, wday_type) %>%
    summarize(steps_avg = mean(steps))

# verify results
# https://www.statmethods.net/stats/descriptives.html
head(steps_avg_wday)
names(steps_avg_wday)
summary(steps_avg_wday)

# part 2: create time series facet plot
# source: http://zevross.com/blog/2019/04/02/easy-multi-panel-plots-in-r-using-facet_wrap-and-facet_grid-from-ggplot2/
install.packages("ggplot2")
library(ggplot2)
ggplot(data=steps_avg_wday, aes(interval, steps_avg)) +
    geom_line(color="blue") +
    geom_point(color="blue") + 
    facet_wrap(~ wday_type)

```